*Supplementary Material 1*

### 2022-02-26 ms/manuscript.Rmd
Add bootstrap tree

tar_load(
  c(
    accs_exclude,
    fern_sanger_seqs_raw,
    fossil_calibration_tips,
    monophy_by_clade,
    plastid_calibration_dates,
    plastid_tree_dated,
    plastome_alignment,
    plastome_alignment_tbl,
    plastome_metadata_raw_all,
    plastome_metadata_renamed,
    plastome_metadata_renamed,
    plastome_seqs_combined_filtered,
    plastome_tree,
    ppgi_taxonomy,
    pteridocat,
    raw_fasta_all,
    sanger_accessions_selection,
    sanger_alignment,
    sanger_alignment_tbl,
    sanger_con_tree_rooted,
    sanger_seqs_combined_filtered,
    sanger_seqs_rogues_removed,
    ts_fossil_calibration_tips
  ),
  store = ftol_cache
)

# Targets from MS workflow
tar_load(
  c(
    accepted_species,
    sanger_sampling,
    coverage_by_rank,
    total_coverage,
    bb_coverage_by_rank,
    bb_total_coverage,
    gb_species_by_year,
    ref_files,
    other_dates,
    crown_ages,
    fossil_ferns_raw,
    family_stem_ages,
    fern_monophy_summ_tbl,
    fern_nonmono_gen,
    plastome_iqtree_log,
    treepl_cv_results
  ),
  store = here::here("_targets")
)
```

@@

```{r phylogeny-stats}
# Bootstrap values for selected clades
ferns_bs <- get_bs(plastome_tree, c("Equisetum_arvense", "Danaea_nodosa"))
hym_gle_bs <- get_bs(plastome_tree, c("Dipteris_conjugata", "Hymenophyllum_polyanthos"))
cyath_salv_bs <- get_bs(plastome_tree, c("Plagiogyria_adnata", "Azolla_filiculoides"))
denn_poly_bs <- get_bs(plastome_tree, c("Dennstaedtia_hirsuta", "Pseudophegopteris_aurita"))

# Mean bootstrap for consensus tree
tibble(
  bs = sanger_con_tree_rooted$node.label)

@@

# Double check that non-monophyletic genera outside of
# top three non-monophyletic subfamilies all have less
# than four genera each
fern_nonmono_gen %>%
  anti_join(
    top_nonmono_subfam,
    by = c("family", "subfamily")) %>%
  count(family, subfamily) %>%
  verify(
    all(n < 4), 
    error_fun = error_msg(
      "Other subfamilies besides top three don't have less than four genera each"),
    success_fun = success_logical)
```

The best-scoring model according to BIC selected for the backbone phylogeny was `r extract_iqtree_mod(plastome_iqtree_log)`.
The tree search completed after `r extract_iqtree_iter(plastome_iqtree_log)` iterations, at which point the bootstrap correlation coefficient of split occurrence frequencies was `r extract_iqtree_corr(plastome_iqtree_log)`.

Ferns are strongly supported as monophyletic (BS `r ferns_bs`; all subsequent relationships mentioned received BS >=`r percent(.98)` in the backbone phylogeny unless otherwise indicated).
The first split within ferns separates a clade including Equisetidae and Ophioglossidae from all other species (`r figure("backbone")`).
The next split separates Marattiales from the remaining ferns, the leptosporangiates (Polypodiidae).
Within leptosporangiate ferns, Osmundales is sister to all other species.
The next lineage to diverge is a clade including Hymenophyllales and Gleicheniales (BS `r hym_gle_bs`), followed by Schizaeales.
Salviniales was recovered as sister to Cyatheales but without strong support (BS `r cyath_salv_bs`), which are in turn sister to Polypodiales. 
The first split within Polypodiales separates a clade including Saccolomatineae and Lindsaeineae from the remainder of species, followed by the subsequent divergences of Pteridineae, then Dennstaedtiineae, which is sister to the eupolypods with moderate support (BS `r denn_poly_bs`).
There are two major clades within eupolypods, Polypodiineae (eupolypods I) and Aspleniineae (eupolypods II).

@@

Bootstrap support was generally across the tree (mean BS) and particularly high at deeper nodes.
Relationships within some more recently diverged clades were less well-supported, including *Cyathea*, *Amauropelta*, and parts of *Dryopteris* and *Elaphoglossum* (`r s_figure("bs")`).

@@

plastome_tree_rooted <- phytools::reroot(
  plastome_tree,
  getMRCA(
    plastome_tree,
    c("Physcomitrium_patens", "Marchantia_polymorpha", "Anthoceros_angustus")
  )) %>%
  ladderize()

@@

# Make tibble for annotating major clades
clade_tibble <- 
  # Use assess_monophy to get MRCA of each major clade
  # They should all by monophyletic anyways
  select(sanger_sampling, species, major_clade) %>%
  assess_monophy(
    taxon_sampling = .,
    tree = plastid_tree_dated,
    tax_levels = "major_clade"
  ) %>%
  get_result_monophy(1) %>%
  # Exclude lycophytes
  filter(!taxon %in% c(
    "Selaginellales", "Lycopodiales", "Isoëtales", "Isoetales")) %>%
  verify(all(monophyly == "Yes")) %>%
  transmute(major_clade = taxon, mrca = parse_number(mrca)) %>%
  left_join(clade_cols_tib, by = "major_clade")
### plot-prep end
```

@@

a <- ggtree(
  plastid_tree_dated, layout = "circular", right = FALSE, size = 0.1) + 
  ftol_geom_cladelab(
    clade_tibble$code[clade_tibble$code %in% c("Eq", "Op", "Ma", "Os", "Gl")], 10) +
  ftol_geom_cladelab(
    clade_tibble$code[clade_tibble$code %in% c("Hy", "Sc", "Sal")], 14) +
  ftol_geom_cladelab(
    clade_tibble$code[clade_tibble$code %in% c("Cy", "Sac", "Li", "De")], 17) +
  ftol_geom_cladelab("Pt", 26) +
  ftol_geom_cladelab("As", 34) +
  ftol_geom_cladelab("Po", 17) +
  scale_color_manual(values = clade_cols) +
  theme(legend.position = "none")

@@

# Trim tree to exemplar species, rename by family
phy <- keep.tip(plastome_tree_rooted, family_exemp$species)
new_tip_label <-
  tibble(species = phy$tip.label) %>%
  left_join(family_exemp, by = "species") %>%
  pull(family)
phy$tip.label <- new_tip_label
# Convert BS support values to numeric
phy$node.label <- parse_number(phy$node.label)
# Scale to total height of 100 for easy plotting
phy <- rescale_tree(phy, 100)

# Make tibble for printing BS values
bs_tibble <- tibble(
  # hard-code node ID: internal nodes start after tip nodes,
  # and phy$node.label is in the same order as internal nodes
  node = 1:Nnode(phy) + Ntip(phy), 
  # Don't print BS = 100%
  bootstrap = ifelse(phy$node.label == 100, "", phy$node.label))

@@

set.seed(123)
# Use the ggtree::`%<+%` operator to map the 
# bootstrap values onto the ggtree object
backbone_plot <- 
  ggtree(phy) %<+% bs_tibble +
  # now we can show BS values using geom_text()
  geom_text(aes(label = bootstrap), hjust = -.25, size = 2.5) +
  # add tip labels
  geom_tiplab(aes(label = label)) +
  # highlight major clades
  geom_hilight(
    data = suborder_labels,
    mapping = aes(
      node = mrca,
      fill = major_clade
    ),
    alpha = 0.2,
    align = "right",
    extend = 30,
    gradient.direction = "rt",
    roundrect = TRUE,
    gradient.length.out = 2) +
  scale_fill_manual(values = clade_cols) +
  # label major clades
  geom_cladelab(
    data = suborder_labels,
    mapping = aes(
      node = mrca,
      label = clade_label
    ),
    offset = 28,
    align = TRUE,
    extend = 0.3,
    hjust = 0,
    barcolour = "grey20",
    barsize = 2) +
  # Label other groups
  geom_text_repel(
    data = other_clade_labels %>%
      filter(clade_label == "Polypodiidae\n(leptosporangiates)"),
    aes(x = x, y = y, label = clade_label),
    hjust = 0,
    xlim = c(NA, 10),
    ylim = c(15, NA),
    color = "grey20",
    segment.color = "grey20"
  ) +
  geom_text_repel(
    data = other_clade_labels %>%
      filter(clade_label == "eupolypods"),
    aes(x = x, y = y, label = clade_label),
    hjust = 0,
    xlim = c(NA, 64),
    ylim = c(32, NA),
    color = "grey20",
    segment.color = "grey20"
  ) +
  geom_text_repel(
    data = other_clade_labels %>%
      filter(clade_label == "Polypodiales"),
    aes(x = x, y = y, label = clade_label),
    hjust = 0,
    xlim = c(NA, 48),
    ylim = c(24, NA),
    color = "grey20",
    segment.color = "grey20"
  ) +
  xlim(0, 180) +
  coord_cartesian(clip = "off") +
  theme(legend.position = "none")
```

>>>>

tar_load(
  c(
    accs_exclude,
    fern_sanger_seqs_raw,
    fossil_calibration_tips,
    monophy_by_clade,
    plastid_calibration_dates,
    sanger_tree_dated,
    plastome_alignment,
    plastome_alignment_tbl,
    plastome_metadata_raw_all,
    plastome_metadata_renamed,
    plastome_metadata_renamed,
    plastome_seqs_combined_filtered,
    plastome_tree_rooted,
    ppgi_taxonomy,
    pteridocat,
    raw_fasta_all,
    sanger_accessions_selection,
    sanger_alignment,
    sanger_alignment_tbl,
    sanger_con_tree_rooted,
    sanger_seqs_combined_filtered,
    sanger_seqs_rogues_removed,
    ts_fossil_calibration_tips
  ),
  store = ftol_cache
)

# Targets from MS workflow
tar_load(
  c(
    accepted_species,
    bb_coverage_by_rank,
    bb_total_coverage,
    coverage_by_rank,
    crown_ages,
    family_stem_ages,
    fern_monophy_summ_tbl,
    fern_nonmono_gen,
    fossil_ferns_raw,
    gb_species_by_year,
    other_dates,
    plastome_iqtree_log,
    ref_files,
    sanger_iqtree_log,
    sanger_sampling,
    total_coverage,
    treepl_cv_results
  ),
  store = here::here("_targets")
)
```

@@

```{r phylogeny-stats}
# Bootstrap values for selected clades
ferns_bs <- get_bs(plastome_tree_rooted, c("Equisetum_arvense", "Danaea_nodosa"))
hym_gle_bs <- get_bs(plastome_tree_rooted, c("Dipteris_conjugata", "Hymenophyllum_polyanthos"))
cyath_salv_bs <- get_bs(plastome_tree_rooted, c("Plagiogyria_adnata", "Azolla_filiculoides"))
denn_poly_bs <- get_bs(plastome_tree_rooted, c("Dennstaedtia_hirsuta", "Pseudophegopteris_aurita"))

# Tibble of bootstrap for consensus trees
bs_tib <-
tibble(bs = sanger_con_tree_rooted$node.label, dataset = "sanger") %>%
bind_rows(tibble(bs = plastome_tree_rooted$node.label, dataset = "plastome")) %>%
  mutate(
    bs = na_if(bs, "Root") %>% 
      parse_number
    )

# Percent of plastome phylogeny with 100% BS support
plastome_bs100 <-
bs_tib %>%
  filter(dataset == "plastome") %>%
  filter(!is.na(bs)) %>%
  mutate(is_100 = if_else(bs == 100, "true", "false")) %>%
  tabyl(is_100) %>%
  mutate(percent = percent(percent, accuracy = 0.1)) %>%
  split(.$is_100)

# Mean BS values for printing
bs <-
bs_tib %>%
  group_by(dataset) %>%
  summarize(
    mean = mean(bs, na.rm = TRUE),
    sd = sd(bs, na.rm = TRUE)
  ) %>%
  mutate(across(c(mean, sd), ~number(., accuracy = 0.1))) %>%
  split(.$dataset)

@@

# Double check that non-monophyletic genera outside of
# top three non-monophyletic subfamilies all have less
# than four genera each
fern_nonmono_gen %>%
  anti_join(
    top_nonmono_subfam,
    by = c("family", "subfamily")) %>%
  count(family, subfamily) %>%
  verify(
    all(n < 4), 
    error_fun = error_msg(
      "Other subfamilies besides top three don't have less than four genera each"),
    success_fun = success_logical)

# Check that model for plastome and sanger was same
assertthat::assert_that(
  extract_iqtree_mod(sanger_iqtree_log) == 
    extract_iqtree_mod(plastome_iqtree_log),
  msg = "Same model not selected for plastome and sanger trees"
)
```

The `r extract_iqtree_mod(plastome_iqtree_log)` model was selected according to BIC for both the plastome (backbone) and Sanger phylogenies.
The tree search completed after `r extract_iqtree_iter(plastome_iqtree_log)` and `r extract_iqtree_iter(sanger_iqtree_log)` iterations, at which point the bootstrap correlation coefficient of split occurrence frequencies was `r extract_iqtree_corr(plastome_iqtree_log)` and `r extract_iqtree_corr(sanger_iqtree_log)` for the plastome and Sanger phylogenies, respectively.

Ferns are strongly supported as monophyletic (BS `r ferns_bs`; all subsequent relationships mentioned received BS >=`r percent(.98)` in the backbone phylogeny unless otherwise indicated).
The first split within ferns separates a clade including Equisetidae and Ophioglossidae from all other species (Figures `r figure_num("backbone")`, `r s_figure_num("backbone_full")`).
The next split separates Marattiales from the remaining ferns, the leptosporangiates (Polypodiidae).
Within leptosporangiate ferns, Osmundales is sister to all other species.
The next lineage to diverge is a clade including Hymenophyllales and Gleicheniales (BS `r hym_gle_bs`), followed by Schizaeales.
Salviniales was recovered as sister to Cyatheales but without strong support (BS `r cyath_salv_bs`), which are in turn sister to Polypodiales. 
The first split within Polypodiales separates a clade including Saccolomatineae and Lindsaeineae from the remainder of species, followed by the subsequent divergences of Pteridineae, then Dennstaedtiineae, which is sister to the eupolypods with moderate support (BS `r denn_poly_bs`).
There are two major clades within eupolypods, Polypodiineae (eupolypods I) and Aspleniineae (eupolypods II).

@@

Bootstrap support was generally across the tree (mean `r bs$sanger$mean` ± `r bs$sanger$sd`; Sanger phylogeny) and particularly high at deeper nodes (mean `r bs$plastome$mean` ± `r bs$plastome$sd`; `r plastome_bs100$true$percent` of nodes with 100% BS; backbone phylogeny).
Relationships within some genera were less well-supported, including *Cyathea*, *Amauropelta*, and parts of *Dryopteris* and *Elaphoglossum* (`r s_figure("bs")`).

@@

@@

# Make tibble for annotating major clades
clade_tibble <- 
  # Use assess_monophy to get MRCA of each major clade
  # They should all by monophyletic anyways
  select(sanger_sampling, species, major_clade) %>%
  assess_monophy(
    taxon_sampling = .,
    tree = sanger_tree_dated,
    tax_levels = "major_clade"
  ) %>%
  get_result_monophy(1) %>%
  # Exclude lycophytes
  filter(!taxon %in% c(
    "Selaginellales", "Lycopodiales", "Isoëtales", "Isoetales")) %>%
  verify(all(monophyly == "Yes")) %>%
  transmute(major_clade = taxon, mrca = parse_number(mrca)) %>%
  left_join(clade_cols_tib, by = "major_clade")
### plot-prep end
```

@@

a <- ggtree(
  sanger_tree_dated, layout = "circular", right = FALSE, size = 0.1) + 
  ftol_geom_cladelab(
    clade_tibble$code[clade_tibble$code %in% c("Eq", "Op", "Ma", "Os", "Gl")], 10) +
  ftol_geom_cladelab(
    clade_tibble$code[clade_tibble$code %in% c("Hy", "Sc", "Sal")], 14) +
  ftol_geom_cladelab(
    clade_tibble$code[clade_tibble$code %in% c("Cy", "Sac", "Li", "De")], 17) +
  ftol_geom_cladelab("Pt", 26) +
  ftol_geom_cladelab("As", 34) +
  ftol_geom_cladelab("Po", 17) +
  scale_color_manual(values = clade_cols) +
  theme(legend.position = "none")

@@

# Trim tree to exemplar species, rename by family
phy <- keep.tip(plastome_tree_rooted, family_exemp$species)
new_tip_label <-
  tibble(species = phy$tip.label) %>%
  left_join(family_exemp, by = "species") %>%
  pull(family)
phy$tip.label <- new_tip_label
# Convert BS support values to numeric
# phy$node.label <- parse_number(phy$node.label)
# Scale to total height of 100 for easy plotting
phy <- rescale_tree(phy, 100)

# Make tibble for printing BS values
bs_tibble <- make_bs_tbl(phy) %>%
  # only print values < 100
  mutate(bs = case_when(
    bs == 100 ~ "",
    TRUE ~ as.character(bs)
  ))

@@

set.seed(123)
# Use the ggtree::`%<+%` operator to map the 
# bootstrap values onto the ggtree object
backbone_plot <- 
  ggtree(phy) %<+% bs_tibble +
  # now we can show BS values using geom_text()
  geom_text(aes(label = bs), hjust = -.25, size = 2.5) +
  # add tip labels
  geom_tiplab(aes(label = label)) +
  # highlight major clades
  geom_hilight(
    data = suborder_labels,
    mapping = aes(
      node = mrca,
      fill = major_clade
    ),
    alpha = 0.2,
    align = "right",
    extend = 30,
    gradient.direction = "rt",
    roundrect = TRUE,
    gradient.length.out = 2) +
  scale_fill_manual(values = clade_cols) +
  # label major clades
  geom_cladelab(
    data = suborder_labels,
    mapping = aes(
      node = mrca,
      label = clade_label
    ),
    offset = 28,
    align = TRUE,
    extend = 0.3,
    hjust = 0,
    barcolour = "grey20",
    barsize = 2) +
  # Label other groups
  geom_text_repel(
    data = other_clade_labels %>%
      filter(clade_label == "Polypodiidae\n(leptosporangiates)"),
    aes(x = x, y = y, label = clade_label),
    hjust = 0,
    xlim = c(NA, 10),
    ylim = c(15, NA),
    color = "grey20",
    segment.color = "grey20"
  ) +
  geom_text_repel(
    data = other_clade_labels %>%
      filter(clade_label == "eupolypods"),
    aes(x = x, y = y, label = clade_label),
    hjust = 0,
    xlim = c(NA, 64),
    ylim = c(32, NA),
    color = "grey20",
    segment.color = "grey20"
  ) +
  geom_text_repel(
    data = other_clade_labels %>%
      filter(clade_label == "Polypodiales"),
    aes(x = x, y = y, label = clade_label),
    hjust = 0,
    xlim = c(NA, 48),
    ylim = c(24, NA),
    color = "grey20",
    segment.color = "grey20"
  ) +
  xlim(0, 180) +
  coord_cartesian(clip = "off") +
  theme(legend.position = "none")
```

### 2022-02-26 ms/si.Rmd
Add bootstrap tree

tar_load(
  c(
    sanger_alignment_tbl,
    sanger_sampling,
    sanger_accessions_selection,
    plastome_tree,
    plastid_tree_dated,
    fossil_calibration_tips,
    ppgi_taxonomy,
    monophy_by_clade,
    acc_table_long,
    plastome_seqs_combined_filtered,
    sanger_tree_rooted,
    fossil_calibrations_for_treepl,
    sanger_tree_dated,
    fossil_ferns_all),
  store = ftol_cache
)

# Targets from MS workflow
tar_load(
  c(
    accepted_species,
    gb_species_by_year,
    ref_files,
    other_dates,
    du_dates_all,
    family_stem_ages,
    pteridocat_ppgi_diff_refs_out,
    pteridocat_ppgi_diff_notes,
    fern_monophy_summ_tbl,
    fern_nonmono_gen,
    spermato_trees
  ),
  store = here::here("_targets")
)
```

@@

>>>>

tar_load(
  c(
    acc_table_long,
    fossil_calibration_tips,
    fossil_calibrations_for_treepl,
    fossil_ferns_all,
    monophy_by_clade,
    plastome_seqs_combined_filtered,
    plastome_tree,
    plastome_tree_rooted,
    ppgi_taxonomy,
    sanger_accessions_selection,
    sanger_con_tree_rooted,
    sanger_sampling,
    sanger_tree_dated,
    sanger_tree_rooted,
    sanger_alignment_tbl
  ),
  store = ftol_cache
)

# Targets from MS workflow
tar_load(
  c(
    du_dates_all,
    family_stem_ages,
    fern_monophy_summ_tbl,
    fern_nonmono_gen,
    gb_species_by_year,
    other_dates,
    pteridocat_ppgi_diff_notes,
    pteridocat_ppgi_diff_refs_out,
    ref_files,
    spermato_trees,
    accepted_species
  ),
  store = here::here("_targets")
)
```

@@

```{r backbone-full-plot}
phy <- plastome_tree_rooted

# Make tibble for printing BS values
bs_tibble <- make_bs_tbl(phy) %>%
  # only print values < 100
  mutate(bs = case_when(
    bs == 100 ~ "",
    TRUE ~ as.character(bs)
  ))

# Make plot
bb_full_plot <- 
  ggtree(phy, size = 0.2, color = "grey10") %<+% 
  bs_tibble +
  # now we can show BS values using geom_text()
  geom_text(
    aes(label = bs), hjust = -.25, size = 1.0,
    color = "grey50") +
  # add tip labels
  geom_tiplab(aes(label = label), size = 1.0)
```

```{r backbone-full-save, eval = params$doc_type == "doc"}
ggsave(
  plot = bb_full_plot,
  file = (result_file(s_figure("backbone_full"), "pdf")),
  height = 20, width = 8, units = "in",
  limitsize = FALSE)
```

**`r s_figure("bs")`**.
Consensus plastome (backbone) phylogeny (`r result_file(s_figure("backbone_full"), "pdf") %>% fs::path_file()`).
Values at nodes are bootstrap support (%); only values less than 100% are shown.

```{r bs-plot}
phy <- sanger_con_tree_rooted

# Make tibble for printing BS values
bs_tibble <- make_bs_tbl(phy)

# Make plot
# optional: round version
bs_plot_round <- 
  ggtree(
    phy,
    layout = "circular",
    branch.length = "none", 
    size = 0.1, color = "grey30") %<+% bs_tibble +
  # Add node points
  geom_nodepoint(
    aes(color = bs), size = 0.2) +
  scale_color_viridis_c(
    "BS",
    option = "B", na.value = "transparent",
    breaks = 0.25*0:4, labels = percent(0.25*0:4))

# Make plot
bs_plot <- 
  ggtree(
    phy,
    size = 0.1, color = "grey30") %<+% bs_tibble +
  # Add node points
  geom_nodepoint(
    aes(color = bs), size = 0.1) +
  scale_color_viridis_c(
    "BS",
    option = "B", na.value = "transparent",
    breaks = 0.25*0:4, labels = percent(0.25*0:4)) +
  # Add tip labels
  geom_tiplab(aes(label = label), size = 0.5, color = "grey30") +
  theme(legend.position = "bottom")
```

```{r bs-save, eval = params$doc_type == "doc"}
ggsave(
  plot = bs_plot,
  file = (result_file(s_figure("bs"), "pdf")),
  height = 60, width = 8, units = "in",
  limitsize = FALSE)
```

**`r s_figure("bs")`**.
Consensus phylogeny (Sanger dataset) with node labels showing bootstrap support (%) (`r result_file(s_figure("bs"), "pdf") %>% fs::path_file()`).
The backbone (plastome) phylogeny was used as a constraint tree, so all clades present in the backbone phylogeny tree receive 100% bootstrap support.
